---
title: "water_syd_05202025"
author: "Chaoyang Ye"
format: html
editor: visual
execute:
  warning: false
  message: false
---

#### Explore water quality at Sydney beaches related to weather, time of day, date and geographical location

1.  **Examine the water quality data by itself through time**

```{r load packages}

library(tidytuesdayR)
library(ggplot2)
library(tidyr)
library(dplyr)
library(GGally)
library(gridExtra)
library(lubridate)
library(rlang)
library(plotly)
```

```{r load data}
tuesdata <- tidytuesdayR::tt_load('2025-05-20')
water_quality <- tuesdata$water_quality
weather <- tuesdata$weather
```

3 key measurements in the water_quality dataframe: enterococci_cfu_100ml, water_temperature_c, conductivity_ms_cm

remove water_temperature_c \> 100, which is likely caused by error

log transform enterococci_cfu_100ml

check their correlation with scatter plot

```{r water quality vs temperature}
water_quality_tf <- water_quality |> 
  filter(water_temperature_c < 100) |>
  mutate(log10_entero = log10(enterococci_cfu_100ml+1)) |>
  mutate(log10_conduct = log10(conductivity_ms_cm+1))
key_measure <- water_quality_tf |>
  select(log10_entero, log10_conduct, water_temperature_c)
ggpairs(key_measure)
```

Does not appear to have correlations between water temperature and bacteria or conductivity

Next, check out relationship of bacteria and conductivity with time of day and day of the year

```{r water quality vs time of the day}
plot_tod <- function(df, col) {
  ggplot(df, aes(x = time, y = !!sym(col))) +
  geom_point(color = "steelblue", size = 1) +
  labs(title = paste0(col, " Over Time of Day"),
       x = "Time of Day", y = col) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10, face = "bold"),
    axis.title.x = element_text(size = 7),
    axis.title.y = element_text(size = 7),
    axis.text.x = element_text(size = 5),
    axis.text.y = element_text(size = 5)
  )
}

p1 <- plot_tod(water_quality_tf, "log10_entero")
p2 <- plot_tod(water_quality_tf, "log10_conduct")

grid.arrange(p1, p2, ncol = 2)
```

Not seeing particular pattern other than measurements were mostly taken during day time.

```{r water quality vs day of the year}
water_quality_tf <- water_quality_tf |>
  mutate(
    year = year(date),
    doy = yday(date)
  )

plot_doy <- function(df, col) {
  ggplot(df, aes(x = doy, y = !!sym(col), color = factor(year))) +
  geom_point() +
  labs(title = paste0(col, "Over Day of Year (Across Years)"),
       x = "Day of Year", y = "log10_entero", color = "Year") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10, face = "bold"),
    axis.title.x = element_text(size = 7),
    axis.title.y = element_text(size = 7),
    axis.text.x = element_text(size = 5),
    axis.text.y = element_text(size = 5),
    legend.text = element_text(size = 5),
    legend.title = element_text(size = 5)
  )
}

p1 <- plot_doy(water_quality_tf, "log10_entero")
p2 <- plot_doy(water_quality_tf, "log10_conduct")
grid.arrange(p1, p2, ncol = 2)
```

```{r smooth out water quality vs day of the year}
plot_doy_smooth <- function(df, col) {
  ggplot(df, aes(x = doy, y = !!sym(col), color = factor(year))) +
  geom_smooth(method = "loess", span = 0.7, se = FALSE) +
  labs(title = paste0(col, " vs. Day of Year (Across Years)"),
       x = "Day of Year", y = col, color = "Year") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10, face = "bold"),
    axis.title.x = element_text(size = 7),
    axis.title.y = element_text(size = 7),
    axis.text.x = element_text(size = 5),
    axis.text.y = element_text(size = 5),
    legend.text = element_text(size = 5),
    legend.title = element_text(size = 5)
  )
}

p1 <- plot_doy_smooth(water_quality_tf, "log10_entero")
p2 <- plot_doy_smooth(water_quality_tf, "log10_conduct")
grid.arrange(p1, p2, ncol = 2)
```

For most years, bacteria count is the lowest at around day 250 (late winter and early spring), and highest at the end of the year (summer); while conductivity showed the opposite trend for recent years

Now check the bacteria count throughout the years by each region

```{r geographic location impact}
all_loc <- names(table(water_quality_tf$region))

plot_region <- function(df, loc) {
  ggplot(df[df$region == loc, ], aes(x = doy, y = log10_entero, color = factor(year))) +
  geom_smooth(method = "loess", span = 0.7, se = FALSE) +
  labs(title = paste0("Measure vs. Day of Year in ", loc),
       x = "Day of Year", y = "log10_entero", color = "Year") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10, face = "bold"),
    axis.title.x = element_text(size = 7),
    axis.title.y = element_text(size = 7),
    axis.text.x = element_text(size = 5),
    axis.text.y = element_text(size = 5),
    legend.text = element_text(size = 5),
    legend.title = element_text(size = 5)
  )
}

all_plot <- lapply(all_loc, function(x)plot_region(water_quality_tf, x))
do.call(grid.arrange, c(all_plot, ncol = 2))
```

Sydney harbor experiences less changes throughout the year than other regions. Southern Sydney had the widest swing during a year. Calculate yearly swings by region

```{r yearly bacteria count range}
water_entero_range <- water_quality_tf |>
  select(region, year, log10_entero) |>
  group_by(region, year) |>
  summarise(
    max_count = max(log10_entero, na.rm = T),
    min_count = min(log10_entero, na.rm = T),
    count_diff = max_count - min_count,
    .groups = "drop"
  )

ggplot(water_entero_range, aes(x = year, y = count_diff, color = region)) +
  geom_line() +
  labs(title = paste0("log10_entero Yearly Swing by Region Over Years"),
       x = "Year", y = "max(log10_entero) - min(log10_entero)") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10, face = "bold"),
    axis.title.x = element_text(size = 7),
    axis.title.y = element_text(size = 7),
    axis.text.x = element_text(size = 5),
    axis.text.y = element_text(size = 5),
    legend.text = element_text(size = 5),
    legend.title = element_text(size = 5)
  )
```

2.  **Explore the relationship of water quality with weather**

    since there is no granular geological information about the weather, only date, temp and precipitation, I will ignore the latitude and longitude column and merge it with the water quality table by date.

```{r merge weather with water quality data}
water_quality_weather <- merge(
  water_quality_tf[, c("region", "date", "time", "water_temperature_c", "log10_entero", "log10_conduct", "year", "doy")], 
  weather[, c("date", "max_temp_C", "min_temp_C", "precipitation_mm")],
  by.x = "date",
  by.y = "date"
  )

water_quality_weather <- water_quality_weather |>
  mutate(temp_diff = max_temp_C - min_temp_C)

key_measure <- water_quality_weather |>
  select(log10_entero, log10_conduct, max_temp_C, min_temp_C, temp_diff, precipitation_mm)
ggpairs(key_measure)
```

Obvious correlation between min_temp_C and max_temp_C, max_temp_C and temp_diff. There seems to be a negative correlation between temp_diff and precipitation_mm, and between log10_entero and temp_diff and precipitation_mm. Does that mean less temperature swing during a day and/or low precipitation is correlated with higher bacteria count? What happens on days of low temp_diff [and]{.underline} low precipitation_mm?

```{r 3D plot to examine relations among bacteria count, temp_diff and precipitation}
water_quality_weather |>
  select(log10_entero, temp_diff, precipitation_mm, year, region) |>
  plot_ly(x = ~log10_entero, 
          y = ~temp_diff, 
          z = ~precipitation_mm,
          color = ~region,
          type = "scatter3d", 
          mode = "markers", 
          size = 2, 
          alpha = 0.6)
water_quality_weather |>
  select(log10_entero, temp_diff, precipitation_mm, year, region) |>
  plot_ly(x = ~log10_entero, 
          y = ~temp_diff, 
          z = ~precipitation_mm,
          color = ~year,
          type = "scatter3d", 
          mode = "markers", 
          size = 2, 
          alpha = 0.6)
```

Highest bacteria counts came from Sydney harbor, and the counts were highest pre-2000

Next we can examine temp_diff and precipitation_mm across the year

```{r temperature difference and precipitation over the years}
p1 <- plot_doy_smooth(water_quality_weather, "temp_diff")
p2 <- plot_doy_smooth(water_quality_weather, "precipitation_mm")
grid.arrange(p1, p2, ncol = 2)
```

Temperature differences are increasing during late winter time in recent years, explaining the lower bacteria counts. Precipitation is decreasing during the same period, also contributing to lower bacteria counts.
